---
title: "Effectiveness of Vitamiin C Delivery Methods on Tooth Growth"
author: "Cha Li"
date: "06/20/2015"
output: pdf_document
---

# Introduction
The purpose of this analysis is to examine the effect of vitamin C on tooth growth in guinea pigs.
There are two control variables in the ToothGrowth dataset: delivery method and dosage. This report
will focus on the effectiveness of the two delivery methods at varying dosage levels.

```{r, echo=T, message=F, warning=FALSE}
# load useful libraries and the dataset
library(datasets)
library(dplyr)
library(ggplot2)

data(ToothGrowth)
```

# Exploratory Analysis
The graphy below groups the treatments by dosage and then within each dosage level the data is split
again by delivery method. In the .5mg and 1mg scenarios we can see that there appears to be a difference
in tooth growth between the delivery methods. This does not seem to be the case, however, in the
2mg scenario.  

```{r}
# create a new column which is just, the other two factors combined. this is 
# for easy graphing
working_data <- ToothGrowth %>%
    mutate(treatment=interaction(ToothGrowth$supp, ToothGrowth$dose)
           , dose=as.factor(dose)
           , supp=as.factor(supp)) %>%
    rename(method=supp, dosage=dose, growth=len)

ggplot(working_data) +
    geom_boxplot(aes(y=growth, x=method, fill=dosage)) +
    geom_point(aes(y=growth, x=method)) + 
    facet_grid(.~dosage, ) +
    ggtitle("Tooth Growth by Dosage and Delivery Method") +
    labs(x="Delivery Method", y="Tooth Growth")
```

We can also see here that the variances between delivery method within each dosage is very different.
This is an important observation that will be used later on for building confidence intervals.
```{r}
summary_data <- working_data %>%
    group_by(dosage, method) %>%
    summarize(n=n(), mu_growth=mean(growth), var_growth=var(growth))
print(summary_data)
```

# Parameters and Assumptions
Before we build confidence intervals and perform hypothesis tests, we make the following assumptions:

1. Although the same 10 guinea pigs were used, we will assume that the trials between delivery 
methods for each dosage are independent. The dataset description implied some amount of time
went by between treatments. 

2. The sample sizes are small for each group so we'll assume a T-distribution.

3. The variance between delivery methods given a dosage value is very different.


# Confidence Intervals and Hypothesis Testing
Using the three points above, the 95% confidence intervals will be defined assuming a two-sample 
unpooled t-distribution. The confidence interval will be built using the following equation:
$$(\hat{X}_1 - \hat{X}_2) \pm t_{\alpha/2, n+m-2} \times \sqrt{\frac{s_1^2}{n} + \frac{s_2^2}{m}}$$
$$n = m = 10, \alpha = .05, df = 18$$
The critical result from defining the interval is when 0 is outside the interval. If this occurs we 
can reject the null hypothesis, $H_0: \mu_1 = \mu_2$, in favor of the alternate hypothesis, 
$H_A: \mu_1 \neq \mu_2$.

```{r}
t_score <- qt(.975, 18)

# get the means and variances we need for the first dosage level (.5,g)
dosage <- summary_data %>% filter(dosage==0.5)
delta <- dosage$mu_growth[1] -dosage$mu_growth[2]

# the confidence interval
ci_dosage1 <- delta + c(-1, 1) * t_score * sqrt(
    ((dosage$var_growth[1]/dosage$n[1])) + (dosage$var_growth[2]/dosage$n[2]))

# repeat the process for the 1mg and 2mg
dosage <- summary_data %>% filter(dosage==1)
delta <- dosage$mu_growth[1] -dosage$mu_growth[2]
ci_dosage2 <- delta + c(-1, 1) * t_score * sqrt(
    ((dosage$var_growth[1]/dosage$n[1])) + (dosage$var_growth[2]/dosage$n[2]))

dosage <- summary_data %>% filter(dosage==2)
delta <- dosage$mu_growth[1] -dosage$mu_growth[2]
ci_dosage3 <- delta + c(-1, 1) * t_score * sqrt(
    ((dosage$var_growth[1]/dosage$n[1])) + (dosage$var_growth[2]/dosage$n[2]))

results <-rbind(ci_dosage1, ci_dosage2, ci_dosage3)
```

# Results
```{r, echo=FALSE}
print(results)
```
The calculated confidence intervals show more precisely that there are significant differences
in tooth growth between the delivery methods with .5mg (dosage1) and 1mg (dosage2) dosages and 
that there is no significant difference with the 2mg dosage. Essentially, we reject the null 
hypothesis in the first two cases but fail to reject it in the third case.





