---
title: "Gas Mileage Regression Analysis"
author: "Coursera: Regression Models"
date: "06/21/2014"
output: pdf_document
---

## Executive Summary

This report will analyse the data provided in the Motor Trend dataset (\texttt{mtcars}), found here <http://bit.ly/1svZI4M>, and
provide some insight into the following two research questions:

1. Is an automatic or manual transmission better for MPG?
2. How different is the MPG between automatic and manual transmissions?

Linear regression models and other techniques will be applied to the data to explore and quantify any differences between
automatic and manual trasmission vehcles.

## Exploratory Analysis
The \texttt{mtcars} dataset contains 32 rows (instances) and 11 columns (attributes). Each instance represents a vehicle type and the attributes describe 
various properties of the vehicles. This analysis will focus specifically on \texttt{mpg} (miles per gallon) and \texttt{am} (transmission type) which are continuous and binary categorical attributes, respectively. \texttt{am} takes on the following values: \texttt{0 = automatic, 1 = manual}. \texttt{mpg} is the response variable and \texttt{am} is the explanatory variable. Of the 32 vehicles, 19 of them have automatic transmissions and 13 have manual transmissions.
The boxplot in the appendix shows the spread of the \texttt{mpg} variable with respect to each transmission type. It is clear from the boxplot that 
some difference exists and that it may be significant. Upon first glance we can see that the medians of both groups are quite far from each other and the IQR's
of each transmission type do not overlap. This means the middle 50% of each group is complete separated from each other. Additionally there do not seem to be any 
outliers at this point in the analysis. We can also see that the variance of the \texttt{automatic} group is smaller than the variance of the \texttt{manual}
group. In the next section we will walk through a regression analysis and formally quantify the differences between the two groups.

## Regression Analysis
As stated earlier, \texttt{mpg} is the response variable and \texttt{am} is the explanatory varible. We will begin by looking at the most basic model first and then consider interactions and confounding variables. \texttt{am = 'automatic'} is the reference level so the linear model we use is:
$$ Y_i = \beta_0 + I(X_i = manual)\beta_1 + \beta_2 \cdot X_i + I(X_i = manual) \cdot \beta_3 \cdot X_i $$
where $I()$ is an indicator function which equals 1 if the condition is met.

Analysis 1 in the appendix shows the summary of the model based on the equation above. \textbf{The expected \texttt{mpg} of an automatic transmission vehicle is $17.7$ and the expected \texttt{mpg} for a manual transmission vehicle is $24.4$.}. Figure 2 shows the Q-Q plot of the residuals and they are near-normally 
distributed as they should be. At first glance it seems like driving a manual transmission vehicle gives you an extra 
$7$ miles per gallon, this is great news, however, the finer details of the model suggest a deeper investigation is needed. The adjusted $R^2$ value is a low 
$.3385$ meaning the model does not do a good job explaining the variance seen in \texttt{mpg}. The rest of this report will consider adjusted $R^2$ values over
standard $R^2$ as a metric for model selection since the former penalizes complex models and the latter increases monotonically

\newpage

## Confounding Variables and Model Selection
There are $9$ variables in the \texttt{mtcars} dataset, each of which may have an impact on \texttt{mpg}. I know from experience that the weight of a car
and the horsepower of an engine both decrease gas mileage as they increase. The \texttt{mtcars} dataset abbreviates these two variables as \texttt{wt} and
\texttt{hp}, respectively. Figure 3 in the appendix plots the distributions of weight and horsepower against transmission type. In both cases we can see that
vehicles in this dataset with a automatic transmission also tend to have larger weights and higher horse power. This raises the obvious question: Do these
two variables have a larger impact on gas mileage than transmission type? In an effort to keep this report short, I am going to assume the other variables 
have no effect.

Analysis 2 and 3 in the appendix aims to answer the new question. We fit three models. The base model is the model we have used up to this point and the two
successive models each add one additional variable, weight followed by horse power. Two methods are used to evaluate if these two variables should be 
added, the first (Analysis 2) is based on \texttt{anova} and the second (Analysis) is based on adjusted $R^2$ values. The p-values (Analysis 2) of the successive
models in the \texttt{anova} suggest that the new variables are significant at the 95% significance level. Reinforcing this result is Analysis 3 which shows
the adjusted $R^2$ value increasing significantly for each new variable, more than doubling when \texttt{wt} is included.

## Conclusion

Analysis 4 shows the summary of the final model which includes \texttt{am}, \texttt{wt}, and \texttt{hp} as regressors of \texttt{mpg}. The adjusted $R^2$ is 
much better than the original model, $.823$ versus $.338$ and we get additional insights into the true effect of transmission type on gas mileage. Interpreting
the properties of the final model with respect to the orignal research questions, we arrive at the following conclusion: \textbf{ Holding weight and horsepower
constant, switching to a manual transmission increases gas mileage by 2 mpg when compared to an automatic transmission. The p-value is only significant at the 
85\% confidence level so this may not be the strongest result.} These results may be different if we did not make the assumption that all other variables
had no effect on mpg but recall that this assumption was made to avoid a long analysis. Only a couple examples were needed to demonstrate the role of confounding
variables in model selection.

\newpage

## Appendix

```{r setup}
# setup the dataset
data(mtcars)
mtcars$am[mtcars$am == '0'] = 'automatic'
mtcars$am[mtcars$am == '1'] = 'manual'
groups = split(mtcars, mtcars$am)
```

```{r}
# ----- Figure 1
boxplot(mpg ~ am, data=mtcars, col=c('red', 'blue'), ylab='MPG', xlab='Transmission Type', main='Transmission Type and Mileage')
summary(groups$automatic$mpg) # summary of mpg for automatic transmission vehicles
summary(groups$manual$mpg) # summary of mpg for manual transmission vehicles
```

\newpage

```{r}
# basic linear regression with mpg as the response and am as the regressor
fit = lm(mpg ~ as.factor(am), data=mtcars)

# ----- Analysis 1
summary(fit)

# ----- Figure 2
qqnorm(fit$residuals); qqline(fit$residuals)
```

\newpage
```{r}
# ----- Figure 3
par(mfrow=c(1, 2))
boxplot(wt ~ am, data=mtcars, col=c('red', 'blue'), ylab='Weight (lb/1000)', xlab='Transmission Type', main='Transmission Type and Weight')
boxplot(hp ~ am, data=mtcars, col=c('red', 'blue'), ylab='Horsepower', xlab='Transmission Type', main='Transmission Type and Horsepower')

fit1 = lm(mpg ~ as.factor(am), data=mtcars)
fit2 = lm(mpg ~ as.factor(am) + wt, data=mtcars)
fit3 = lm(mpg ~ as.factor(am) + wt + hp, data=mtcars)

# ----- Analysis 2
anova(fit1, fit2, fit3)

# ----- Analysis 3
# Adjusted R^2 values for each successive model from above
print(c(summary(fit1)$adj.r.squared, summary(fit2)$adj.r.squared, summary(fit3)$adj.r.squared))

# ----- Final Analysis 4
summary(fit3)
```



